Comparing probabilistic predictive models applied to football

### 2015-08-14 template.tex
final remarks update

    Bothered by such concerns, our goal was to answer the following question: is it possible, using a simple probabilistic model, to predict results of football matches better than more sophisticated or complex models?
    To answer this, we compared two sophisticated models that consider multiple covariates with simple multinomial-Dirichlet models that consider only the number of matches won, tied or lost by each team as inputs.
    The models were tested for the second round of the Brazilian football championships of first division and compared using standard metrics or scoring rules.
    We also used other criteria such as the proportion of matches that were predicted ``correctly'' by each model and a measure of calibration, both explained in detail below.
    According to the adopted criteria, the simple models showed a similar predictive performance to the complex ones.

    The predictions of the two benchmark models were published on Internet websites before every matchday, and are treated by us as black-box models (BB1 and BB2).
    Since one of the authors is responsible for one of these models and the other was described in a dissertation \citep{arruda2000}, we know that both models are based on Holgate bivariate Poisson regression considering as explanatory variables the home field advantage, attack and defense strength for each team, and provide the probabilities of all possible outcomes (home team wins, draws or loses) for each match.
    Because both models are based on elaborate parametric assumptions, they are suitable for the comparison we are proposing: simple versus complex models. To be precise, we would have to define what is a ``simple'' model and what is a ``complex'' model. Based on the opening quotations, one may say that the number of parameters defines the complexity of a model, which is not exactly the definition we will use, since we consider other criteria, like the interpretation of explanatory variables and mathematical structure of the statistical model.

@@

    This paper is organized as follows.
    Section \ref{sec::experimental} describes the proposed models, Section \ref{sec::scoring} presents the scoring rules and criteria used in this work to classify the models.
    Section \ref{sec::results} reports the performance of the models predicting 190 matches comprising the second round of the Brazilian championship of first division--each team plays 19 matches each round, since the championship has 20 teams.
    In Section \ref{sec::remarks} we discuss the results and close with proposals for future research.

@@

    The benchmark models assume that the number of goals $(Y_1, Y2)$ scored respectively by teams A (home team) and B (away
    team) follows a bivariate Poisson distribution \citep{Holgate64} with parameters $(\lambda_1, \lambda_2, \lambda_3)$,
    which has p.m.f given by
\begin{equation*}
P(Y_1 = y_1, Y_2 = y_2 | \lambda_1, \lambda_2, \lambda_3) =
\exp\{-(\lambda_1 + \lambda_2 + \lambda_3)\}
 \sum_{k = 0}^{\min(y_1, y_2)} \dfrac{\lambda_1^{x - k} \lambda_2^{y - k} \lambda_3^k}{(x-k)!(y -
 k)!k!}, \label{eq::pois.biv}
\end{equation*}
for $\lambda_1, \lambda_2 > 0$ and $\lambda_3 \geq 0$.

Both marginal distributions of $(Y_1, Y_2)$ follow Poisson
distributions with dependence parameter $\lambda_3 \geq 0$. If
$\lambda_3 = 0$ the marginal distributions are independent, while if
$\lambda_3 > 0$ the marginal distributions are positively
correlated. Because of its flexibility, \cite{Karlis2003} argues
that this distribution is a plausible choice for modeling dependence
of scores in sports competitions.

@@

Similar to \cite{Lee97} and \cite{Karlis2003}, both benchmark models
adopt the following log-linear link functions
\begin{align*}
\log(\lambda_1) &= \mu + \text{ATT}_A - \text{DEF}_B + \gamma, \\
\log(\lambda_2) &= \mu + \text{ATT}_B - \text{DEF}_A,
\end{align*}
where $\mu$ is a parameter representing the average number of goals
in a match, $\text{ATT}_k$ is the offensive strength of team k,
$\text{DEF}_k$ is the defensive strength of team k and $\gamma$ is
the home advantage for team A, $k = A, B$.

The prediction for an upcoming match-day is obtained by fitting the
model to all the observed scores up to the previous match-day and
then summing up the probabilities of all scores relevant to the win,
draw and loss outcomes of future matches.

@@

        Before we explain and illustrate the proposed models with an example, we make two further remarks.
        The first one is that we will separately consider home and away games for each team. This will allow us to take into account the different performances under those conditions.
        The second remark is that the prediction of an upcoming match between two teams $A$ and $B$ can be done using
        the past performance of either team $A$ or $B$.
        To solve this problem, a compromise between the two predictive distributions was adopted.
        The next subsections \ref{sec::Mn_Dir1}, \ref{sec::Mn_Dir2} and \ref{sec::Mn_Dir3} present three different possible solutions.

@@

    According to standard scoring rules and other criteria here adopted to qualify the predictions, our findings show that the predictive power of the three simple models was quite similar to the benchmark models.
    %We compared the probabilistic predictions of three simple models with the ones provided by two more complex models for the second round of the 2014 Brazilian football championships of first and second divisions.
    This work thus indicates that while there has been a boom in the literature with the proposal of several fancy and involved models, especially in the era of big data, simple models can many times do
    as well as their complex counterparts.

    This is often unnoticed by users of such methods, which leads to the overuse of models that are
    hard to interpret and to implement, and that also require one to keep track of several covariates. We would call this phenomenon the {\it king's nudity} in allusion to Anderson's {\it The Emperor's New Clothes}, \cite{emperor}: ``But he isn't wearing anything at all!''. However, this is not always the case. In this spirit, since we have explored only the prediction of football matches, we would like to study if the same extends to other fields such as, for instance, weather or economics forecasts.

>>>>

    Bothered by such concerns, our goal was to answer the following question: is it possible, using a simple probabilistic model, to predict results of football matches better than more sophisticated or complex models?
    To answer this, we compared two sophisticated models that consider multiple covariates with simple multinomial-Dirichlet models that consider only the number of matches won, tied or lost by each team as inputs.
    The models were tested for the second round of the Brazilian football championship of first division and compared using standard metrics or scoring rules.
    We also used other criteria such as the proportion of matches that were predicted ``correctly'' by each model and a measure of calibration, both explained in detail below.
    According to the adopted criteria, the simple models showed a similar predictive performance to the complex ones.

    The predictions of the two benchmark models were published on Internet websites before every matchday, and are treated by us as black-box models (BB1 and BB2).
    Since one of the authors is responsible for one of these models and the other was described in a dissertation \citep{arruda2000}, we know that both models are based on Holgate bivariate Poisson regression considering as explanatory variables the home field advantage, attack and defense strength of each team, and provide the probabilities of all possible outcomes (home team wins, draws or loses) for each match.
    Because both models are based on elaborate parametric assumptions, they are suitable for the comparison we are proposing: simple versus complex models.

    To be precise, we would have to define what is a ``simple'' model and what is a ``complex'' model. Based on the opening quotations, one may say that the number of parameters defines the complexity of a model, which is not exactly the definition we will use, since we consider other criteria, like the interpretation of explanatory variables and mathematical structure of the statistical model.

@@

    This paper is organized as follows.
    Section \ref{sec::experimental} describes the proposed models, Section \ref{sec::scoring} presents the scoring rules and criteria used in this work to classify the models.
    Section \ref{sec::results} reports the performance of the models predicting 190 matches comprising the second round of the Brazilian championship of first division---each team plays 19 matches each round, since the championship has 20 teams.
    In Section \ref{sec::remarks} we discuss the results and close with proposals for future research.

@@

    The benchmark models assume that the number of goals $(Y_1, Y_2)$ scored respectively by teams $A$ (home team) and $B$ (away
    team) has a bivariate Poisson distribution \citep{Holgate64} with parameters $(\lambda_1, \lambda_2, \lambda_3)$,
    which has probability mass function given by
\begin{equation*}
P(Y_1 = y_1, Y_2 = y_2 | \lambda_1, \lambda_2, \lambda_3) =
\exp\{-(\lambda_1 + \lambda_2 + \lambda_3)\}
 \sum_{k = 0}^{\min(y_1, y_2)} \dfrac{\lambda_1^{x - k} \lambda_2^{y - k} \lambda_3^k}{(x-k)!(y -
 k)!k!}, \label{eq::pois.biv}
\end{equation*}
for $\lambda_1, \lambda_2 > 0$ and $\lambda_3 \geq 0$.

Both marginal distributions of $(Y_1, Y_2)$ has Poisson
distributions with dependence parameter $\lambda_3 \geq 0$. If
$\lambda_3 = 0$ the marginal distributions are independent, while if
$\lambda_3 > 0$ the marginal distributions are positively
correlated. Because of its flexibility, \cite{Karlis2003} argues
that this distribution is a plausible choice for modeling dependence
of scores in sports competitions.

@@

Similar to \cite{Lee97} and \cite{Karlis2003}, both benchmark models
adopt the following log-linear link functions
\begin{align*}
\log(\lambda_1) &= \mu + \text{ATT}_A - \text{DEF}_B + \gamma, \\
\log(\lambda_2) &= \mu + \text{ATT}_B - \text{DEF}_A,
\end{align*}
where $\mu$ is a parameter representing the average number of goals
in a match, $\text{ATT}_k$ is the offensive strength of team $k$,
$\text{DEF}_k$ is the defensive strength of team $k$ and $\gamma$ is
the home advantage, $k = A, B$.

The prediction for an upcoming matchday is obtained by fitting the
model to all the observed scores up to the previous match-day and
then summing up the probabilities of all scores relevant to the win,
draw and loss outcomes of future matches.

@@

        Before we explain and illustrate the proposed models with an example, we make two further remarks.
        The first one is that we will separately consider home and away games for each team, allowing us to take into account the different performances under these conditions.
        The second remark is that the prediction of an upcoming match between two teams $A$ and $B$ can be done using
        the past performance of either team $A$ or $B$.
        To solve this problem, a compromise between the two predictive distributions was adopted.
        The next subsections \ref{sec::Mn_Dir1}, \ref{sec::Mn_Dir2} and \ref{sec::Mn_Dir3} present three different possible solutions.

@@

    According to standard scoring rules and other criteria here adopted to qualify the predictions, our findings show that the predictive power of the three simple models was quite similar to the benchmark models.
    %We compared the probabilistic predictions of three simple models with the ones provided by two more complex models for the second round of the 2014 Brazilian football championships of first and second divisions.
    {\color{red}This work thus indicates that while there has been a boom in the literature with the proposal of several fancy and involved models, especially in the era of big data, simple models can many times predict the final outcome as well as their complex counterparts. Nevertheless, the complex models cited here are able to answer different questions besides the final outcome that the simple multinomial models are not, for instance, they can estimate teams' performance parameters allowing for ranking teams according to attack and deffensive quality. Thus, we can not assert that complex models are useless.}

    This is often unnoticed by users of such methods, which leads to the overuse of models that are
    hard to interpret and to implement, and that also require one to keep track of several covariates. We would call this phenomenon the {\it king's nudity} { in allusion to Anderson's {\it The Emperor's New Clothes}, \cite{emperor}: ``But he isn't wearing anything at all!''. However, this is not always the case. In this spirit, since we have explored only the prediction of football matches, we would like to study if the same extends to other fields such as, for instance, weather or economics forecasts.
